Component response rate variation drives stability in large complex systems

### 2018-07-24 notebook/SI_science.tex
Last commit before shifting to natcom

In a synthesis of eco-evolutionary feedbacks on community stability,
Patel et al. model a system that includes a vector of potentially
changing species densities (\(\mathbf{N}\)) and a vector of potentially
evolving traits
(\(\mathbf{x}\))\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}.
For any species \(i\) or trait \(j\), change in species density
(\(N_{i}\)) or trait value (\(x_{j}\)) with time (\(t\)) is a function
of the vectors \(\mathbf{N}\) and \(\mathbf{x}\),

@@

In the above, \(f_{i}\) and \(g_{j}\) are functions that define the
effects of all species densities and trait values on the density of a
species \(i\) and the value of trait \(j\), respectively. Patel et al.
were interested in stability when the evolution of traits was relatively
slow or fast in comparison with the change in species
densities\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}, and
this is modulated in the above by the scalar \(\epsilon\). The value of
\(\epsilon\) thereby determines the timescale separation between ecology
and evolution, with high \(\epsilon\) modelling relatively fast
evolution and low \(\epsilon\) modelling relative slow
evolution\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}.

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following
May\textsuperscript{\protect\hyperlink{ref-May1972}{2},\protect\hyperlink{ref-May1973}{3}},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(m_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{3}},

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{\gamma} \mathbf{M}\) modulates the values of components and
can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{2},\protect\hyperlink{ref-May1973}{3}}.

@@

\includegraphics{SI_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

    \NormalTok{C_dat  <-}\StringTok{ }\KeywordTok{rbinom}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1000} \NormalTok{*}\StringTok{ }\DecValTok{1000}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{, }\DataTypeTok{prob =} \DecValTok{1}\NormalTok{);}

@@

@@

    \NormalTok{A1vec       <-}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(A1_vm);}

@@

    \KeywordTok{plot}\NormalTok{(A0_r, A0_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{16.5}\NormalTok{, }\FloatTok{15.5}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{16.5}\NormalTok{,}\FloatTok{15.5}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }
         \DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{, }\DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{cex.lab =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{cex.axis =} \FloatTok{1.5}\NormalTok{, }
         \DataTypeTok{asp =} \DecValTok{1}\NormalTok{);}

@@

    \NormalTok{x1 <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\DecValTok{1000}\NormalTok{) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec) *}\StringTok{ }\KeywordTok{cos}\NormalTok{(vl) +}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{diag}\NormalTok{(A1));}
    \NormalTok{y1 <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\DecValTok{1000}\NormalTok{) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec) *}\StringTok{ }\KeywordTok{sin}\NormalTok{(vl);}

@@

    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{x1, }\DataTypeTok{y =} \NormalTok{y1, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{lty =} \StringTok{"dashed"}\NormalTok{);}
    \KeywordTok{plot}\NormalTok{(A1_r, A1_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{16.5}\NormalTok{, }\FloatTok{15.5}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{16.5}\NormalTok{,}\FloatTok{15.5}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{,}

@@

    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{x1, }\DataTypeTok{y =} \NormalTok{y1, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{)}
    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{x0, }\DataTypeTok{y =} \NormalTok{y0, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{lty =} \StringTok{"dashed"}\NormalTok{);}

@@

\includegraphics{SI_files/figure-latex/unnamed-chunk-9-1.pdf}

@@

While the foundational work of
May\textsuperscript{\protect\hyperlink{ref-May1972}{2}} applies broadly
to complex networks, much attention has been given specifically to
ecological networks of interacting species. In these networks, the
matrix \(M\) is interpreted as a community matrix and each row and
column is interpreted as a single species. The effect that the density
of any species \(i\) has on the population dynamics of species \(j\) is
found in \(M_{ij}\), meaning that \(M\) holds the effects of pair-wise
interactions between \(S\)
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{4}--\protect\hyperlink{ref-Tang2014b}{6}}.
While May's original
work\textsuperscript{\protect\hyperlink{ref-May1972}{2}} considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(M\) are negative), mutualist networks
(all off-diagonal elements of \(M\) are positive), and predator-prey
networks (for any pair of \(i\) and \(j\), the effect of \(i\) on \(j\)
is negative and \(j\) on \(i\) is positive, or vice
versa)\textsuperscript{\protect\hyperlink{ref-Allesina2012}{4}--\protect\hyperlink{ref-Allesina2011}{7}}.
In general, competitor and mutualist networks tend to be unstable, while
predator-prey networks tend to be highly stabilising.

I investigated competitor, mutualist, and predator-prey networks
following Allesina et
al.\textsuperscript{\protect\hyperlink{ref-Allesina2012}{4}}. To create
these networks, I first generated a random matrix \(M\), then changed
the elements of \(M\) accordingly. If \(M\) was a competitive network,
then the sign of any positive off-diagonal elements was reversed to be
negative. If \(M\) was a mutualist network, then the sign of any
positive off-diagonal elements was reversed to be positive. And if \(M\)
was a predator-prey network, then all \(i\) and \(j\) pairs of elements
were checked; any pairs of the same sign were changed so that one was
negative and the other was positive. The \texttt{species\_interaction}
function used to do this is below.

@@

Overall, as
expected\textsuperscript{\protect\hyperlink{ref-Allesina2012}{4}},
predator-prey communities are relatively stable while mutualist
communties are highly unstable. But interestingly, while \(Var(\gamma)\)
stabilises predator-prey and competitor communities, it does not
stabilise mutualist communities. This is unsurprising because purely
mutualist communities are characterised by a very
positive\textsuperscript{\protect\hyperlink{ref-Allesina2012}{4}}
leading \(\Re(\lambda)\), and it is highly unlikely that \(Var(\gamma)\)
alone will shift all real parts of eigenvalues to negative values.

@@

\includegraphics{SI_files/figure-latex/unnamed-chunk-24-1.pdf}

@@

Ideally, to investigate the potential of \(Var(\gamma)\) for increasing
the proportion of stable complex systems, the search space of all
possible \(\gamma\) vectors would be evaluated for each unique \(M\).
This is technically impossible because \(\gamma_{i}\) can take any real
value between 0-2, but even rounding \(\gamma\) to reasonable values
would result in a search space too large to practically explore. Under
these conditions, genetic algorithms are highly useful tools for finding
practical solutions by mimicking the process of biological
evolution\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{8}}. In
this case, the practical solution is finding vectors of \(\gamma\) that
decrease the most positive real eigenvalue of \(M\). The genetic
algorithm below achieves this by initialising a large population of 1000
different potential \(\gamma\) vectors and allowing this population to
evolve through a process of mutation, crossover (swaping \(\gamma_{i}\)
values between vectors), selection, and reproduction until either a
\(\gamma\) vector is found where all \(\Re(\lambda) < 0\) or some
``giving up'' critiera is met (in the below, this ``giving up''"
criteria is met when 20 generations pass, or if the fitness increase
from one generation to the next is below a certain criteria). The
genetic algorithm relies on five functions. The first outer function
\texttt{Evo\_rand\_gen\_var} runs all of the simulations
(\texttt{max\_sp} refers to the maximum \(S\) value simulated, and
\texttt{iters} refers to the number of \(M\) to try for each \(S\)).

@@

\includegraphics{SI_files/figure-latex/unnamed-chunk-32-1.pdf}

@@

For complex systems in which individual system components (\(S\))
represent the density of some tangible quantity, it is important to
consider the feasibility of the system. Feasibile equilibria assume that
the values of all system components are positive at
equilibrium\textsuperscript{\protect\hyperlink{ref-Grilli2017}{9}--\protect\hyperlink{ref-Song2018}{11}}.
This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically
realistic\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{10}}.
Consequently, the use of random matrices and traditional stability
critiera for making inferences in theoretical analyses of species
networks has recently been
criticised\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{10}}.
While the key results in the main text are intended to be general to all
complex systems, and not restricted to species networks, I have also
performed a feasibility analysis on all matrices \(M\). This analysis
reveals that feasibility is not affected by \(Var(\gamma)\), meaning
that for pure interacting species networks, variation in component
response time (i.e., species generation time) does not affect stability
at biologically realistic species densities. Nevertheless, ecological
interactions do not exist in isolation in empirical systems, but instead
interact with
evolutionary\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}},
abiotic, or social-economic systems. The relevance of \(\gamma\) for
complex system stability presented in the main text should therefore not
be ignored in the broader context of ecological communities.

Dougoud et al.\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{10}}
define the following feasibility criteria for ecological systems
characterised by \(S\) interacting species with varying densities.

@@

In the above, \(x^{*}\) is the vector of species abundances at
equilibrium (for feasibility, all values in \(x^{*}\) must be positive).
The matrix \(I\) is the identity matrix (1s on the diagonal, 0s on the
off-diagonal elements), and the value \(\theta\) is strength of
intraspecific competition (diagonal elements). As I have done elsewhere,
diagonal values are set to \(-1\), so \(\theta = -1\). The variable
\(C\) is the inter-connectivity (i.e., `connectance') of the community,
which was set to \(C = 1\) throughout the manuscript and supplemental
information, except \protect\hyperlink{connectance}{where otherwise
noted}. The variable \(\delta\) is a normalisation parameter that
modulates the strength of interactions (\(\sigma\) in the main text),
which are held in \(A\). In the main text, implicitly, \(\delta = 0\)
underlying strong interactions. Hence, the whole \((CS)^{-\delta} = 1\),
so in the above, a diagonal matrix of -1s (\(\theta I\)) is added to
\(A\), which has a diagonal of all zeros and an off-diagonal affecting
species interactions (i.e., the expression \((CS)^{-\delta}\) relates to
May's\textsuperscript{\protect\hyperlink{ref-May1972}{2}} stability
criterion\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{10}} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < -1\), and hence
\((CS)^{-\delta} = 1\) for the randomly simulated systems in the main
text and supplemental information). The above criteria is therefore
reduced to the below; note that the parenthetical in both equations
produces an \(M\) matrix as used throughout the main text and
supplemental information,

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-Patel2018}{}
1. Patel, S., Cortez, M. H. \& Schreiber, S. J. Partitioning the effects
of eco-evolutionary feedbacks on community stability. \emph{American
Naturalist} \textbf{191,} 1--29 (2018).

\hypertarget{ref-May1972}{}
2. May, R. M. Will a large complex system be stable? \emph{Nature}
\textbf{238,} 413--414 (1972).

\hypertarget{ref-May1973}{}
3. May, R. M. Qualitative stability in model ecosystems. \emph{Ecology}
\textbf{54,} 638--641 (1973).

\hypertarget{ref-Allesina2012}{}
4. Allesina, S. \& Tang, S. Stability criteria for complex ecosystems.
\emph{Nature} \textbf{483,} 205--208 (2012).

\hypertarget{ref-Allesina2015a}{}
5. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

\hypertarget{ref-Tang2014b}{}
6. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\hypertarget{ref-Allesina2011}{}
7. Allesina, S. \& Levine, J. M. A competitive network theory of species
diversity. \emph{Proceedings of the National Academy of Sciences of the
United States of America} \textbf{108,} 5638--5642 (2011).

\hypertarget{ref-Hamblin2013}{}
8. Hamblin, S. On the practical usage of genetic algorithms in ecology
and evolution. \emph{Methods in Ecology and Evolution} \textbf{4,}
184--194 (2013).

\hypertarget{ref-Grilli2017}{}
9. Grilli, J. \emph{et al.} Feasibility and coexistence of large
ecological communities. \emph{Nature Communications} \textbf{8,} (2017).

\hypertarget{ref-Dougoud2018}{}
10. Dougoud, M., Vinckenbosch, L., Rohr, R., Bersier, L.-F. \& Mazza, C.
The feasibility of equilibria in large ecosystems: a primary but
neglected concept in the complexity-stability debate. \emph{PLOS
Computational Biology} \textbf{14,} e1005988 (2018).

\hypertarget{ref-Song2018}{}
11. Song, C. \& Saavedra, S. Will a small randomly assembled community
be feasible and stable? \emph{Ecology} \textbf{99,} 743--751 (2018).

>>>>

In a synthesis of eco-evolutionary feedbacks on community stability,
Patel et al. model a system that includes a vector of potentially
changing species densities (\(\mathbf{N}\)) and a vector of potentially
evolving traits (\(\mathbf{x}\)) (15). For any species \(i\) or trait
\(j\), change in species density (\(N_{i}\)) or trait value (\(x_{j}\))
with time (\(t\)) is a function of the vectors \(\mathbf{N}\) and
\(\mathbf{x}\),

@@

In the above, \(f_{i}\) and \(g_{j}\) are functions that define the
effects of all species densities and trait values on the density of a
species \(i\) and the value of trait \(j\), respectively. Patel et al.
were interested in stability when the evolution of traits was relatively
slow or fast in comparison with the change in species densities (15),
and this is modulated in the above by the scalar \(\epsilon\). The value
of \(\epsilon\) thereby determines the timescale separation between
ecology and evolution, with high \(\epsilon\) modelling relatively fast
evolution and low \(\epsilon\) modelling relative slow evolution (15).

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following May (1, 19), the value of a component \(i\) at
time \(t\) (\(v_{i}(t)\)) is affected by the value of \(j\)
(\(v_{j}(t)\)) and \(j\)'s marginal effect on \(i\) (\(m_{ij}\)), and by
\(i\)'s response rate (\(\gamma_{i}\)),

@@

In matrix notation (20),

@@

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{\gamma} \mathbf{M}\) modulates the values of components and
can be analysed using the techniques of May (1, 19).

@@

\includegraphics{SI_science_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

    \NormalTok{C_dat  <-}\StringTok{ }\KeywordTok{rbinom}\NormalTok{(}\DataTypeTok{n =} \DecValTok{1000} \NormalTok{*}\StringTok{ }\DecValTok{1000}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{, }\DataTypeTok{prob =} \FloatTok{0.05}\NormalTok{);}

@@

    \NormalTok{A_mat  <-}\StringTok{ }\KeywordTok{species_interactions}\NormalTok{(A_mat, }\DataTypeTok{type =} \DecValTok{0}\NormalTok{);}

@@

@@

    \KeywordTok{plot}\NormalTok{(A0_r, A0_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{6}\NormalTok{, }\FloatTok{2.25}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{5}\NormalTok{,}\FloatTok{4.25}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{,}

@@

@@

    \KeywordTok{plot}\NormalTok{(A1_r, A1_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{6}\NormalTok{, }\FloatTok{2.25}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{5}\NormalTok{,}\FloatTok{4.25}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{,}

@@

    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{x0, }\DataTypeTok{y =} \NormalTok{y0, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{lty =} \StringTok{"solid"}\NormalTok{);}

@@

\includegraphics{SI_science_files/figure-latex/unnamed-chunk-9-1.pdf}

@@

While the foundational work of May (1) applies broadly to complex
networks, much attention has been given specifically to ecological
networks of interacting species. In these networks, the matrix \(M\) is
interpreted as a community matrix and each row and column is interpreted
as a single species. The effect that the density of any species \(i\)
has on the population dynamics of species \(j\) is found in \(M_{ij}\),
meaning that \(M\) holds the effects of pair-wise interactions between
\(S\) species (2, 4, 20). While May's original work (1) considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(M\) are negative), mutualist networks
(all off-diagonal elements of \(M\) are positive), and predator-prey
networks (for any pair of \(i\) and \(j\), the effect of \(i\) on \(j\)
is negative and \(j\) on \(i\) is positive, or vice versa) (2, 4, 17,
20). In general, competitor and mutualist networks tend to be unstable,
while predator-prey networks tend to be highly stabilising.

I investigated competitor, mutualist, and predator-prey networks
following Allesina et al. (2). To create these networks, I first
generated a random matrix \(M\), then changed the elements of \(M\)
accordingly. If \(M\) was a competitive network, then the sign of any
positive off-diagonal elements was reversed to be negative. If \(M\) was
a mutualist network, then the sign of any positive off-diagonal elements
was reversed to be positive. And if \(M\) was a predator-prey network,
then all \(i\) and \(j\) pairs of elements were checked; any pairs of
the same sign were changed so that one was negative and the other was
positive. The \texttt{species\_interaction} function used to do this is
below.

@@

Overall, as expected (2), predator-prey communities are relatively
stable while mutualist communties are highly unstable. But
interestingly, while \(Var(\gamma)\) stabilises predator-prey and
competitor communities, it does not stabilise mutualist communities.
This is unsurprising because purely mutualist communities are
characterised by a very positive (2) leading \(\Re(\lambda)\), and it is
highly unlikely that \(Var(\gamma)\) alone will shift all real parts of
eigenvalues to negative values.

@@

\includegraphics{SI_science_files/figure-latex/unnamed-chunk-24-1.pdf}

@@

Ideally, to investigate the potential of \(Var(\gamma)\) for increasing
the proportion of stable complex systems, the search space of all
possible \(\gamma\) vectors would be evaluated for each unique \(M\).
This is technically impossible because \(\gamma_{i}\) can take any real
value between 0-2, but even rounding \(\gamma\) to reasonable values
would result in a search space too large to practically explore. Under
these conditions, genetic algorithms are highly useful tools for finding
practical solutions by mimicking the process of biological evolution
(16). In this case, the practical solution is finding vectors of
\(\gamma\) that decrease the most positive real eigenvalue of \(M\). The
genetic algorithm below achieves this by initialising a large population
of 1000 different potential \(\gamma\) vectors and allowing this
population to evolve through a process of mutation, crossover (swaping
\(\gamma_{i}\) values between vectors), selection, and reproduction
until either a \(\gamma\) vector is found where all \(\Re(\lambda) < 0\)
or some ``giving up'' critiera is met (in the below, this ``giving up''"
criteria is met when 20 generations pass, or if the fitness increase
from one generation to the next is below a certain criteria). The
genetic algorithm relies on five functions. The first outer function
\texttt{Evo\_rand\_gen\_var} runs all of the simulations
(\texttt{max\_sp} refers to the maximum \(S\) value simulated, and
\texttt{iters} refers to the number of \(M\) to try for each \(S\)).

@@

\includegraphics{SI_science_files/figure-latex/unnamed-chunk-32-1.pdf}

@@

For complex systems in which individual system components (\(S\))
represent the density of some tangible quantity, it is important to
consider the feasibility of the system. Feasibile equilibria assume that
the values of all system components are positive at equilibrium (5, 18,
21). This is of particular interest for ecological communities because
population density cannot take negative values, meaning that ecological
systems need to be feasible for stability to be biologically realistic
(18). Consequently, the use of random matrices and traditional stability
critiera for making inferences in theoretical analyses of species
networks has recently been criticised (18). While the key results in the
main text are intended to be general to all complex systems, and not
restricted to species networks, I have also performed a feasibility
analysis on all matrices \(M\). This analysis reveals that feasibility
is not affected by \(Var(\gamma)\), meaning that for pure interacting
species networks, variation in component response time (i.e., species
generation time) does not affect stability at biologically realistic
species densities. Nevertheless, ecological interactions do not exist in
isolation in empirical systems, but instead interact with evolutionary
(15), abiotic, or social-economic systems. The relevance of \(\gamma\)
for complex system stability presented in the main text should therefore
not be ignored in the broader context of ecological communities.

Dougoud et al. (18) define the following feasibility criteria for
ecological systems characterised by \(S\) interacting species with
varying densities.

@@

In the above, \(x^{*}\) is the vector of species abundances at
equilibrium (for feasibility, all values in \(x^{*}\) must be positive).
The matrix \(I\) is the identity matrix (1s on the diagonal, 0s on the
off-diagonal elements), and the value \(\theta\) is strength of
intraspecific competition (diagonal elements). As I have done elsewhere,
diagonal values are set to \(-1\), so \(\theta = -1\). The variable
\(C\) is the connectance of the community, which was set to \(C = 1\)
throughout the manuscript and supplemental information, except
\protect\hyperlink{connectance}{where otherwise noted}. The variable
\(\delta\) is a normalisation parameter that modulates the strength of
interactions (\(\sigma\) in the main text), which are held in \(A\). In
the main text, implicitly, \(\delta = 0\) underlying strong
interactions. Hence, the whole \((CS)^{-\delta} = 1\), so in the above,
a diagonal matrix of -1s (\(\theta I\)) is added to \(A\), which has a
diagonal of all zeros and an off-diagonal affecting species interactions
(i.e., the expression \((CS)^{-\delta}\) relates to May's (1) stability
criterion (19) by \(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < -1\), and
hence \((CS)^{-\delta} = 1\) for the randomly simulated systems in the
main text and supplemental information). The above criteria is therefore
reduced to the below; note that the parenthetical in both equations
produces an \(M\) matrix as used throughout the main text and
supplemental information,

@@

### 2018-07-24 notebook/ms_for_review.Rmd
Last commit before shifting to natcom

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b) [see also ]. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b). The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.
